Methods for Analyzing Models of\\ Rod-Shaped Bacteria

### 2025-07-29 paper/main.tex
relabel (sub)sections

% --------------------------------------------------------------------------------------------------
\subsection{Visualization}
\label{subsubsection:visualization}

@@

% --------------------------------------------------------------------------------------------------
\subsection{Data Extraction}
\label{section:fitting-and-comparing-masks}

@@

In order to determine if our extraction algorithm provides reliable results, we have to take
advantage of a functionality which will be introduced later in
section~\ref{subsubsection:visualization}.
Our methodology makes use of the fact that we have already set up our model and solving scheme and
are able to produce cell masks from these numerical results.
In doing so, we have exact knowledge about the position of our agents since we can simply obtain
this information from our model.
But on the other hand we can also try to extract this position from the generated cell-masks.
By comparing the differences between them, we get a good estimate of how well our extraction
algorithm can determine the true position of the bacteria.

@@

\paragraph{Comparison across Generations}
In order to compare a variable amount of cells, we can not utilize the positions of these cells
for comparison directly since one particular state of our simulation might contain cells which are
not present in the experimental results (or vice-versa).
We thus need additional methods which can be used for comparisons across division events.
To achieve this, we use the numerical results of the \ac{abm} simulation and our visualization
algorithms to generate cell-masks (see section~\ref{subsubsection:visualization}).
We then compare the generated masks to the experimentally obtained ones pixel by pixel.
Whenever two pixel colors are not matching, a loss value is assigned.
The sum of all those penalties is then the total cost.
This algorithm is an implementation of a signed area measure.
It is very important to take into account the parental relationship between the cells, checking if
either one of the cells is a daughter of the other cell.
Figure~\ref{fig:mask-difference-metric} (A,B) shows two successive states of a numerical simulation
with varying amounts of cells.
We can see that some division events have occurred but other cells have only grown without
completing a division step.
The resulting penalties without accounting for the parental relationships are displayed in
Figure~\ref{fig:mask-difference-metric} (C).
Black indicates that the pixels are matching and no loss is assigned while a white pixel
indicates that the original pixels are not matching, thus assigning a loss value.
Similarly, Figure~\ref{fig:mask-difference-metric} (D) takes the parental relationship into account
and highlights these in gray.
It is possible to assign a custom value for the parental loss value.
If we set this loss value to $0$, this algorithm ensures the contiuity of the loss function in time.

@@

% --------------------------------------------------------------------------------------------------
\subsection{Non-Individual Treatment}
% --------------------------------------------------------------------------------------------------
\subsection{Individual Treatment}
\label{subsec:parameter-estimation-individual-treatment}
\begin{figure}[H]
    \centering
    \includegraphics[width=0.33\textwidth]
        {figures/crm_estimate_params/IWF-Goettingen/rod-lengths-individual.pdf}%
    \includegraphics[width=0.33\textwidth]
        {figures/crm_estimate_params/IWF-Goettingen/rod-lengths-average.pdf}%
    \includegraphics[width=0.33\textwidth]
        {figures/crm_estimate_params/IWF-Goettingen/parameter-distribution.pdf}%
    \caption{}
    \label{fig:estimated-growth-rates}
\end{figure}
% --------------------------------------------------------------------------------------------------
\subsection{With Division}
We will now focus on using the presented methods for parameter estimation via microscopic images.
Our initial state is given by TODO~\cite{}.% TODO
We use our extraction algorithm to obtain the initial positions of our cells.
From there, we supply the model with parameters such that we can make a prediction for the final
positions of the bacteria.
We chose a time-interval which does not contain any division events.
By doing so, we can utilize the simpler and more performant position comparison scheme for our cost
function.
Another important detail is the individuality of the agents.
In our case, this means that each agent might have different parameters associated to it.
For this dataset, we assumed that all interaction parameters which specify the interaction
potential are identical across all agents but growth rates and the physical radii of the cells are
individual and we also optimize them.
We chose to optimize for 3 interaction parameters of the Mie potential $n,m,V_0$ and two parameters
which were individual for each cell, namely the radii $r_i$ and growth rates $\mu_i$.
We also included the damping constant $\lambda$ and removed any stochastic motion contributions
($D=0$).
This leads to a total of $4+2\times 6=16$ parameters.
To estimate these parameters we used a latin-hypercube sweep across the parameter space and
performed a local minimization with the Nelder-Mead method (TODO citations).% CITATIONS

@@

>>>>

% --------------------------------------------------------------------------------------------------
\subsection{Visualization}
\label{subsection:visualization}

@@

% --------------------------------------------------------------------------------------------------
\subsection{Data Extraction}
\label{section:data-extraction}

@@

In order to determine if our extraction algorithm provides reliable results, we have to take
advantage of a functionality which will be introduced later in
section~\ref{subsection:visualization}.
Our methodology makes use of the fact that we have already set up our model and solving scheme and
are able to produce cell masks from these numerical results.
In doing so, we have exact knowledge about the position of our agents since we can simply obtain
this information from our model.
But on the other hand we can also try to extract this position from the generated cell-masks.
By comparing the differences between them, we get a good estimate of how well our extraction
algorithm can determine the true position of the bacteria.

@@

\paragraph{Comparison across Generations}
In order to compare a variable amount of cells, we can not utilize the positions of these cells
for comparison directly since one particular state of our simulation might contain cells which are
not present in the experimental results (or vice-versa).
We thus need additional methods which can be used for comparisons across division events.
To achieve this, we use the numerical results of the \ac{abm} simulation and our visualization
algorithms to generate cell-masks (see section~\ref{subsection:visualization}).
We then compare the generated masks to the experimentally obtained ones pixel by pixel.
Whenever two pixel colors are not matching, a loss value is assigned.
The sum of all those penalties is then the total cost.
This algorithm is an implementation of a signed area measure.
It is very important to take into account the parental relationship between the cells, checking if
either one of the cells is a daughter of the other cell.
Figure~\ref{fig:mask-difference-metric} (A,B) shows two successive states of a numerical simulation
with varying amounts of cells.
We can see that some division events have occurred but other cells have only grown without
completing a division step.
The resulting penalties without accounting for the parental relationships are displayed in
Figure~\ref{fig:mask-difference-metric} (C).
Black indicates that the pixels are matching and no loss is assigned while a white pixel
indicates that the original pixels are not matching, thus assigning a loss value.
Similarly, Figure~\ref{fig:mask-difference-metric} (D) takes the parental relationship into account
and highlights these in gray.
It is possible to assign a custom value for the parental loss value.
If we set this loss value to $0$, this algorithm ensures the contiuity of the loss function in time.

@@

% --------------------------------------------------------------------------------------------------
\subsection{Non-Individual Treatment}
% --------------------------------------------------------------------------------------------------
\subsection{Individual Treatment}
\label{subsec:parameter-estimation-individual-treatment}
\begin{figure}[H]
    \centering
    \includegraphics[width=0.33\textwidth]
        {figures/crm_estimate_params/IWF-Goettingen/rod-lengths-individual.pdf}%
    \includegraphics[width=0.33\textwidth]
        {figures/crm_estimate_params/IWF-Goettingen/rod-lengths-average.pdf}%
    \includegraphics[width=0.33\textwidth]
        {figures/crm_estimate_params/IWF-Goettingen/parameter-distribution.pdf}%
    \caption{}
    \label{fig:estimated-growth-rates}
\end{figure}
% --------------------------------------------------------------------------------------------------
\subsection{Without Division}
We will now focus on using the presented methods for parameter estimation via microscopic images.
Our initial state is given by TODO~\cite{}.% TODO
We use our extraction algorithm to obtain the initial positions of our cells.
From there, we supply the model with parameters such that we can make a prediction for the final
positions of the bacteria.
We chose a time-interval which does not contain any division events.
By doing so, we can utilize the simpler and more performant position comparison scheme for our cost
function.
Another important detail is the individuality of the agents.
In our case, this means that each agent might have different parameters associated to it.
For this dataset, we assumed that all interaction parameters which specify the interaction
potential are identical across all agents but growth rates and the physical radii of the cells are
individual and we also optimize them.
We chose to optimize for 3 interaction parameters of the Mie potential $n,m,V_0$ and two parameters
which were individual for each cell, namely the radii $r_i$ and growth rates $\mu_i$.
We also included the damping constant $\lambda$ and removed any stochastic motion contributions
($D=0$).
This leads to a total of $4+2\times 6=16$ parameters.
To estimate these parameters we used a latin-hypercube sweep across the parameter space and
performed a local minimization with the Nelder-Mead method (TODO citations).% CITATIONS

@@

% --------------------------------------------------------------------------------------------------
\subsection{With Division (TODO)}
